---
title: "Introduction to biostatistics"
subtitle: "Problem 2"
author: "Felix Lenner, Johnny Pellas & Xue Yu"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# Set working directory
knitr::opts_knit$set(root.dir = '~/biostatistics-presentation')
setwd('~/biostatistics-presentation')

options(htmltools.dir.version = FALSE)

library(tidyverse)

data <- read_tsv("data/problem2_data.txt")

data <- data %>%
  mutate(group = factor(group, levels = c('Std', 'Exp')))

```

# Presentation of research question 

> "A randomized trial evaluated a new therapy program designed to help patients recover from a stroke, comparing it to a standard program. Patients were evaluated at baseline (week 0), and for 7 consecutive weeks, using the Barthelindex. This score ranges from 0 to 100, higher scores indicating better functional ability".

---

# Data description 

- How many individuals are there in the data set? 

- Summarize the variables using suitable tables and/or diagrams.

> "The impression is that both groups start at similar levels (how does this relate to randomization?), and gradually increase, but that the experimental group climbs faster. In a linear regression model, this corresponds to two separate slopes, which can be captured by an interaction term."

---

# Statistical analysis

- Which analysis method did you choose? 

- Why? (This should be discussed even if the exam prescribes a certain analysis.) 

- Explain why you think your chosen analysis is reasonable.

- When presenting your results, tell the audience which number(s) you consider most important, and what they mean. Interpret your results in ordinary language.

> "Analyze the data using multiple linear regression with the following predictors: Group, week, and a group x week interaction term. How to include interaction terms depends on the statistical software."

> "You should get 4 estimated regression coefficients, including the intercept. What do they mean? Which one tells us most about the new program? Does it seem promising?"

> "Bonus question: The assumption of independent measurements is violated in this example. Why? Is this a serious problem? Can you think of an alternative analysis that overcomes this problem? What results does it give?"

---

# Conclusions and discussion

- What results did you get?

- How do you interpret them?

- What biological conclusions do you draw?

- Keep the presentation concise. It should take no longer than 15 minutes, followed by approximately 10 minutes discussion.

---

# Example slide

```{r, fig.width=14,fig.height=5, dev='png', dpi=300}

data %>%
  ggplot(aes(week, score, group = week, fill = group)) + 
  geom_boxplot() +
  facet_wrap( ~ group)

```

---

# Example slide

```{r, echo=F, fig.width=14,fig.height=5, dev='png', dpi=300}

data %>%
  ggplot(aes(week, score, group = id, color = factor(id))) +
  geom_line() + 
  facet_wrap(~group)

```

